\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{hyperref}
\usepackage{orcidlink}

\begin{document}
\begin{center}
{\LARGE\bfseries Technology Adoption in Dual Economies: How Informality and Non-Wage Labor Costs Shape Automation Dynamics --- Evidence from Colombia}
\vspace{1cm}

{\large Cristian Espinal Maya\textsuperscript{a,*} \orcidlink{0009-0000-1009-8388}}\\
{\large Santiago Jim\'enez Londo\~no\textsuperscript{a} \orcidlink{0009-0007-9862-7133}}\\
{\large Jovani Alberto Jim\'enez-Builes\textsuperscript{b} \orcidlink{0000-0001-7598-7696}}

\vspace{0.5cm}
\textsuperscript{a} Department of Economics, Universidad EAFIT, Carrera 49 No. 7 Sur-50, Medell\'in 050022, Antioquia, Colombia\\
\textsuperscript{b} Facultad de Minas, Universidad Nacional de Colombia, Medell\'in 050034, Antioquia, Colombia

\vspace{0.5cm}
* Corresponding author: \href{mailto:cjespinalm@eafit.edu.co}{cjespinalm@eafit.edu.co}\\
Tel: +57 305 338 8895
\end{center}

\vspace{1cm}
\section*{CRediT Authorship Contribution Statement}

\textbf{Cristian Espinal Maya:} Conceptualization, Methodology, Software, Data curation, Formal analysis, Writing -- original draft, Visualization.
\textbf{Santiago Jim\'enez Londo\~no:} Conceptualization, Methodology, Validation, Writing -- review \& editing.
\textbf{Jovani Alberto Jim\'enez-Builes:} Supervision, Writing -- review \& editing.

\section*{Acknowledgements}

We thank seminar participants at Universidad EAFIT's Department of Economics for helpful comments. We are grateful to DANE for providing access to the Annual Manufacturing Survey and Gran Encuesta Integrada de Hogares microdata. All errors are our own.

\section*{Declaration and Funding}

The authors declare that they have no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper. This research did not receive any specific grant from funding agencies in the public, commercial, or not-for-profit sectors.

\end{document}
